\chapterwithauthor{Frank Karlitschek}{Underestimating the Value of a Free Software Business Model}

\authorbio{Frank Karlitschek was born in 1973 in Reutlingen, Germany and started to
write software at the age of 11. He studied Computer Science at the University
of T\"ubingen and became involved in free software and Internet technologies in
the mid-1990s. In 2001, he started to contribute to KDE by launching KDE-Look.org,
an artwork community site which later became the openDesktop.org network. Frank
started several Open Source projects and initiatives like the Social Desktop,
the Open Collaboration Services, the Open-PC and ownCloud. In 2007 he started a
company called hive01 which offers services and products around Open Source and
Internet technologies.
Today Frank is a board member and Vice President of the KDE e.V. and a regular
speaker at international conferences.}

\section*{Introduction}

Ten years ago, I underestimated the value of a business model. Free software and
a business model? They do not belong together. At least, that is what I thought
when I started contributing to KDE in 2001. Free Software is about fun and not
money. Right? Free software people want a world where everybody can write
software and huge companies, like Microsoft or Google, are superfluous. Software
should be free and anyone who wants to develop software should be able to do so
-- even hobby developers. So earning money is not important. Right? Today, I
hold a different opinion. Sometimes developers should be remunerated for their
efforts.  

\section*{The Free Software motivation}

Most Free Software developers have two basic motivations to work on Free
Software. The first motivation is the fun factor. It is a fantastic experience
to work together with very talented people from all over the world and create
great technology. KDE, for example, is one of the most welcoming communities I
know. It is so much fun to work with thousands of contributors from all over the
world to create software which will be used by millions. Basically, everyone is
an expert in one or more areas and we collaborate to create a shared vision. For
me it is always a blast to meet other KDE contributors, exchange ideas or work
on our software whether we meet online or in real life at one of the many
conferences or events. And it is also about friendship. Over the years I have
made many good friends in KDE.

But KDE contributors are not motivated only by fun to join KDE. It is also the
idea that all of us can make the world a better place with our contributions. Free
Software is essential if you care about access to technology and IT for
developing countries. It enables poor people to participate in the information
age without buying expensive licenses for proprietary software. It is
essential for people who care about privacy and security, because Free Software
is the only way to see exactly what your computer is doing with your private
data. Free Software is important for a healthy IT eco-system, because it enables
everybody to build on the work of others and really innovate. Without Free
Software it would not have been possible for Google or Facebook to start their
businesses. It is not possible to innovate and create the next disruptive
technology if you depend on proprietary software and do not have full access to
all parts of the software.

Free Software is also essential for education, because everybody can see all the
internals of the software and study how it works. That is how Free Software helps to
make the world a better place and why I contribute to Free Software projects
such as KDE. 

\section*{The need for an ecosystem}

These are the main reasons why I want to see Free Software, and especially the
free desktop, become mainstream. To make this happen, we need a lot more
contributors than we have today. By contributors I mean people who write the
core frameworks, the desktop, the great applications. We need people who work on
usability, artwork, promotion and many other important areas. KDE is already a
really big community with thousands of members. But we need more people to help 
to compete with proprietary software in a big way. The Free Software community
is tiny compared to the proprietary software world. On the one hand this is not
a problem, because the distributed software development model of the Free
Software world is much more efficient than the closed source way of writing
software. One big advantage is, for example, the ability to re-use code better.
But even with these advantages we need many more contributors than we have
today, if we really want to conquer the desktop and mobile markets.

We also need companies to help us bring our work to the mass market. In a
nutshell, we need a big and healthy ecosystem that enables people to work on
Free Software for a living.

\section*{The current situation}

I started contributing to KDE over 10 years ago and since then I have seen
countless highly motivated and talented people join KDE. This is really cool.
The problem is that I also saw a lot of experienced contributors dropping out of
KDE. That is really sad. Sometimes it is just the normal way of the world.
Priorities shift and people concentrate on other stuff. The problem is that many
also drop out because of money. At some point people graduate and want to move
out of their dorm rooms. Later some people want to get married and have kids. At
this point people have to find jobs. There are some companies in the KDE
ecosystem that offer KDE-related jobs. But these are only a fraction of the
available IT jobs. So, a lot of senior KDE contributors have to work for
companies where they work on proprietary software, unrelated to KDE and Free
Software. Sooner or later most of these developers drop out of KDE.
I underestimated this factor 10 years ago, but I think it is a problem for KDE
in the long term, because we lose our most experienced people to proprietary
software companies.

\section*{My dream world}

In my dream world people can pay their rent by working on Free Software and
they can do it in a way which does not conflict with our values.
KDE contributors should have all the time they need to contribute to KDE and
Free Software in general. They should earn money by helping KDE. Their hobbies
should become their jobs. This would make KDE grow in a big way, because it
would be fun to contribute and also provide good long-term job prospects.  

\section*{What are the options?}

So what are the options? What can we do to make this happen? Are there ways for
developers to pay their rent while working on Free Software? I want to list a
few ideas here that I collected during several discussions with Free Software
contributors. Some of them are probably controversial, because they introduce
completely new ideas into the Free Software world. But I think it is essential
for us to think beyond our current world if we want to be successful with our
mission.  

\paragraph*{Sponsored development}

Today, more and more companies appreciate the importance of Free Software and
contribute to Free Software projects, or even release their own completely Free
Software projects. This is an opportunity for Free Software developers. We
should talk to more companies and convince them to work with the Free Software
world. 

\paragraph*{End-user donations}

There should be an easy way for end-users to donate money directly to
developers. If a user of a popular application wants to support the developer
and promote the further development of the application, donating money should be
just one mouse click away. The donation system can be built into the application
to make it as easy as possible to send money.

\paragraph*{Bounties}

The idea behind bounties is that one or more users of an application can pay for
the development of a specific feature. A user can list his feature request on a
website and say how much he is willing to pay for the feature. Other users who
also like the same feature may add some money to the feature request. At some
point the developer starts to develop the feature and collects the money from
the users. This bounty feature is not easy to implement. People already tried to
set up a system like this and failed. But I think it can work if we do it
right. 

\paragraph*{Support}

The idea is that the developer of an application sells direct support to the
users of the application. For example, the users of an application buy support
for, let us say, \$5 a month and get the right to call the developer directly at
specified times of the day, users may post questions to a specific email
address, or the developer can even help the users via a remote desktop. I
realize many developers will not like the idea that users call them and ask
strange questions, but if this means that they earn enough with the support
system to work full-time on their applications, then it must be a good thing.

\paragraph*{Supporters}

This is the idea that end-users can become supporters of an application. The
``Become a Supporter'' button would be directly built into the application. The
user then becomes a supporter for a monthly payment of, for example \$5, which
goes directly to the developer. All the supporters are listed in the About
Dialog of the application together with their photos and real names. Once a year
all supporters are also invited to a special supporter party together with the
developers. It is possible that a developer may be able to work full-time on an
application, if enough users become supporters.

\paragraph*{Affiliate programs}

Some applications have integrated web services and some of these web services
run affiliate programs. For example, a media player can be integrated in the
Amazon mp3 MusicStore or a PDF reader can be integrated in an ebook store.
Every time a user buys content via the application, the developer gets some
money.

\paragraph*{App store for application binaries}

Many people do not know that it is possible to sell binaries of Free Software.
The GPL only requires that you also provide the source code. So, it is perfectly
legal and OK to sell nicely packaged binaries of our software. In fact,
companies such as Red Hat and Novell already sell our software in their commercial distributions but the developers do not benefit from it directly. All the revenue goes to the companies and nothing to the developers. So we could enable the Free Software developers to sell nicely packaged, optimized and
tested applications to the end-user. This might work especially well on Mac or
Windows. I am sure a lot of users would pay \$3 for an Amarok Windows binary, or
digiKam for Mac, if all the money went directly to the developer.

\section*{Conclusion}

Most of these ideas are not easy to implement. They require changes to our
software, changes to our ways of working and changes among our users who must be
encouraged to show they value the software we create by helping to fund its
development.

However, the potential benefits are huge. If we can secure revenue streams for
our software we can retain our best contributors and maybe attract new ones. Our
users will get a better experience with faster software development, the ability
to directly influence development through bounties and better support.

Free Software is no longer just a hobby to be done in your spare time. It is
time to make it a business.
