\chapterwithauthor{Till Jaeger}{On being a Lawyer in FOSS} 

\authorbio{Dr. Till Jaeger has been a partner at JBB Rechtsanwaelte since 2001. He
is a Certified Copyright and Media Law Attorney and advises large and
medium-sized IT businesses as well as government authorities and software
developers on matters involving contracts, licensing and online use. One
particular focus of his work is on the legal issues created by Free and Open
Source Software. He is co-founder of the Institute for Legal Aspects of Free \&
Open Source Software (ifrOSS). He provides advice on compliance with open source licenses and on
compatibility issues, and helps developers and software companies to enforce
licenses. Till represented the gpl-violations.org project in several lawsuits to enforce
the GPL and has published several articles and books related to legal questions
of Free and Open Source Software. He was a member of the Committee C in the
GPLv3 drafting process.}

\noindent{}One thing upfront: I am not a geek. I never have been one, and have no
intention of becoming one in the future. 

Instead, I am a lawyer. Most people who read this book probably tend to
sympathize more with geeks than with lawyers. Nevertheless, I do not want to
hide this fact. That the FOSS community is not necessarily fond of lawyers but
busy developing software is something I \textit{did} know about FOSS in early
1999 when our ways first crossed. But there were also quite a few things I did
not know.

In 1999, while completing my doctoral thesis that focused on a classical
copyright topic, I was assessing the scope of moral rights. In this context I
spent a while pondering about the question of how moral rights of programmers
are safeguarded by the GPL, which allows others to modify their programs. This
is how I first got in contact with FOSS. At the time, ``free'' and ``open''
certainly had different meanings, but the difference was not worth arguing
about in the world I was living in. However, since I was free to do what I was
interested in and open to investigate new copyright questions, I soon found out
that the two words \textit{do} have something in common, that they are
\textit{different} and yet they are best used together...

There are three things I wish I had known back then:

First, my technical knowledge, particularly in the field of software, was
insufficient. Second, I did not really know the community and what mattered to
the people who were part of it. Last but not least, I did not know much about
foreign jurisdictions back then. It would have been useful to know all that from
the beginning.

Since that time, I have learned a fair bit, and just as the community is happy
to share its achievements I am happy to share my lessons\footnote{The ``Institut
f\"ur Rechtsfragen der Freien und Open Source Software'' (Institute for Legal
Questions on Free and Open Source Software) offers, inter alia, a collection of
FOSS related literature and court decisions; see www.ifross.org for details.}:

\paragraph*{Technical knowledge}
How is software architecture shaped? What is the technical structure of software
like? Which licenses are compatible with each other and which are not, and how
and why? How is the Linux kernel structured? 

To name one example, the important question of what constitutes a ``derivative
work'' according to the GPL determines how the software may be licensed.
Everything that counts as derived from GPL-licensed software must be distributed
under the GPL. To assess whether a certain software is a ``derivative work'' or
not requires profound technical understanding. The interaction of program
modules, linking, IPC, plugins, framework technology, header files and so on
determines, among other criteria, whether a program is formally inseparable,
which helps to determine whether it is derived from another program or not. 

\paragraph*{Knowledge of the industry and the community}
Besides these functionality issues I had no profound understanding of the idea
behind FOSS and the motivation of the developers and the companies that use
FOSS. Neither did I really know about its philosophical background, nor was I
familiar with practical issues such as ``who is a maintainer?'' or ``how do version control systems work?'' In order to serve your clients best, these matters are no less important than your proficiency in technical aspects.  
Our clients ask us about legal aspects of forming business models such as dual
licensing, ``open core'', support and services contracts, code development and
code contribution agreements. We consult clients concerning what FOSS might have
in store for their companies or institutions. We also advise developers on what
they can do about infringement of their copyrights, and draft and negotiate
contracts for them. In order to serve such clients comprehensively, it is
important to be familiar with the different points of view.  

\paragraph*{Comparative law knowledge}
The third thing a FOSS lawyer needs is knowledge about foreign jurisdictions, at
least a few, and the more the better. In order to construe the different
licenses, it is essential to be familiar with the perspective of the people who
have drafted it. In most cases the U.S. legal system is of key importance. For
example, the GPL was drafted with U.S. legal concepts in mind. In the United
States, ``distribution'' includes online distribution, whereas under the German
Copyright system there is a distinction between offline and online distribution.
Licenses that have been drafted by lawyers from the United States may thus be
construed as including online distribution, which might be relevant and helpful
in court proceedings\footnote{\url{http://www.ifross.org/Fremdartikel/LGMuenchenUrteil.pdf}, Cf. Welte v. Skype, 2007}.
 
\section*{Always Learning}
So, all this is useful to know. And as software keeps on being developed and
modified to provide solutions for the needs of the day, so my mind will
hopefully keep on finding answers to the challenges the vibrant FOSS community
poses to a lawyer's mind.
