\chapterwithauthor{Jonathan ``Duke'' Leto}{Test-Driven Enlightenment}

\authorbio{Jonathan ``Duke'' Leto is a software developer, published mathematician,
Git ninja and avid bicyclist living in Portland, Oregon. He is a core
developer of Parrot Virtual Machine and founder of Leto Labs LLC.}

\noindent{}When I first got involved in Free and Open Source Software, I had no clue what
tests were or why they were important. I had worked on some personal programming
projects before, but the first time I actually worked on a project with others,
i.e., got a commit bit, was Yacas, Yet Another Computer Algebra System, a computer algebra system similar to Mathematica.

At this stage in my journey, tests were an afterthought. My general
meta-algorithm was to hack on code $\rightarrow$ see if it works $\rightarrow$
write a simple test to show it works (optional). If a test was challenging to
write, it most likely never got written.

This is the first step in the path to Test-Driven Enlightenment. You know tests
are probably a good idea, but you have not seen the benefit of them clearly, so
you only write tests occasionally.

If I could open up a wormhole and tell my younger self one piece of wisdom about
testing, it would be:
\begin{quote}``Some tests, in the long-run, are more important than the code
they test.''\end{quote}

A few people right about now may be thinking that I put on my tinfoil testing
hat when I sit down to code. How can tests be \emph{more} important than the
code they test? Tests are the proof that your code \emph{actually} works, and
they guide you to writing correct code as well as providing the flexibility to
change code and know that features still work. The larger your codebase becomes,
the more valuable your tests are, because they allow you to change one part of
your code and still be sure that the rest of it works.

Another vital reason to write tests is because it indicates that something is
explicitly desirable, rather than an unintended side-effect or oversight. If you
have a specification, you can use tests to verify that you meet it, which is
very valuable, and in some industries, necessary. A test is just like telling a
story, where the story is how you think code should work.

Code either changes and evolves or bitrots\footnote{The term ``bitrot'' is coder
slang for the almost universal fact that if a piece of code does not change but
everything it relies on does, it ``rots'' and usually has very little chance of
working unless modifications are made to accommodate newer software and
hardware.}.

Very often, you will write tests once, but then totally refactor your
implementation or even rewrite it from scratch. Tests often outlive the code
they originally tested, i.e., one set of tests can be used no matter how many
times your code is refactored. Tests are actually the litmus test that allows you to throw away an old implementation and say ``this newer implementation has a much better design and passes our test suite.'' I have seen this happen many
times in the Perl and Parrot communities, where you can often find me.

Tests allow you to change things quickly and know if something is broken. They
are like jet packs for developers.

Carpenters have a bit of sage wisdom that goes like this:
\begin{quote}``Measure twice, cut once.''\end{quote} 

Coding is like cutting and tests are like measuring.

Test-Driven Enlightenment saves an enormous amount of time, because instead of
flailing around, fiddling with code, not having a direction, tests hone your
focus.

Tests also are very good positive feedback. Every time you make a new test pass,
you know that your code is better and it has one more feature or one less bug.

It is easy to think ``I want to add 50 features'' and spend all day fiddling
with code, constantly switching between working on different things. Most of the
time, very little will be accomplished. Test-Driven Enlightenment guides one to
focus on making one test pass at a time.

If you have a single failing test, you are on a mission to make it pass. It focuses
your brain on something very specific, which very often has better results than
switching between tasks constantly.

Most information about being test-driven is very specific to a language or
situation, but that does not need to be the case. Here is how to approach
adding a new feature or fixing a bug in any language:
\begin{enumerate}
 \item Write a test that fails, which you think will pass when the feature is
implemented or bug is fixed. Advanced: As you write the test, run it
occasionally, even if it is not done yet, and guess the actual error message
that the test will give. The more tests you write and run, the easier this will
become.
 \item Hack on the code.
 \item Run the test. If it passes, go to \#4, otherwise go to \#2.
 \item You are done! Do a happy dance :)
\end{enumerate}

This method works for any kind of test and any language. If you remember only one thing about testing from this essay, remember the steps above.

Here are some more general test-driven guidelines that will serve you well and
apply in almost any situation:
\begin{enumerate}
 \item Understand the difference between what is being tested and what is being
used as a tool to test something else.
 \item Fragile tests. You could write a test that makes sure an error message is
exactly correct. But what happens when the error message changes? What happens
when someone internationalizes your code to Catalan? What happens when someone
runs your code on an operating system you have never heard of? The more resilient your test is,
the more valuable it will be.
\end{enumerate}

Think about these things when you write tests. You want them to be resilient,
i.e., tests, for the most part, should only have to change when functionality
changes. If you have to change your tests often, but functionality is not
changing, you are doing something wrong.

\section*{Kinds of tests}

Many people start to get confused when people speak of integration tests,
unit tests, acceptance tests and many other flavors of tests. One
should not worry too much about these terms. The more tests you write, the more
nuances you will see and the differences between tests will become more
apparent. Everyone does not have the same definition for what these tests are,
but the terms are still useful to describe kinds of tests.

\section*{Unit tests vs. integration tests}

Unit tests and integration tests form a spectrum. Unit tests test small bits of
code, and integration tests verify how more than one unit fits together. The test
writer gets to decide what comprises a unit, but most often it is at the level
of a function or method, although some languages call those things by different
names.

To make this a little more concrete, we will give a basic analogy using
functions. Imagine that $f(x)$ and $g(x)$ are two functions which represent two units of code. For
concreteness, let's assume they represent two specific functions in your
favorite Free/Open Source project's codebase.

An integration test asserts something like function composition, i.e., $f(g(a)) =
b$. An integration test is testing how multiple
things integrate or work together, instead of how a single part works
individually. If algebra isn't your thing, another way to look at it is unit
tests only test one part of the machine at a time, but integration tests very
many parts work in unison. A great example of an integration test is test-driving a car.
You are not checking the air pressure, or measuring voltage of the spark plugs.
You are making sure the vehicle works as a whole.

Most of the time it is good to have both. I often start with unit tests and add
integration tests as needed, since you will weed out the most basic bugs first,
then find more subtle bugs that are related to how pieces do not quite fit
together, as opposed to the individual pieces not working. Many people write
integration tests first and then delve into unit tests. Which you write first
is not nearly as important as writing both kinds.

\section*{Enlightenment}

Test-Driven Enlightment is a path, not a place. Enjoy the journey and make sure
to stop and smell the flowers if and when you get lost.
