\chapterwithauthor{Jos Poortvliet}{Big Plans Don't Work}

\authorbio{Jos Poortvliet works as openSUSE community manager for SUSE Linux.
Before that he was active in the international KDE community as team lead for
the marketing team. In his ``offline life'' he has had jobs at a variety of
companies as Business Consultant. His favorite pastime is experimenting in the
kitchen, trying to come up with something edible.}

\begin{quote}``It is better to take many small steps in the right direction than
to make a great leap forward only to stumble backward.'' -- Old Chinese
proverb\end{quote}

\subsection*{A great idea\dots}
Once upon a time in the marketing team of a Free Software project, someone came
up with a great idea to grow the project. A program would be set up to get IT
students to learn about the project and join in. Universities would be contacted
and someone would talk to them to get them interested. Ambassadors would then go
to those universities and give a course there, coaching students in their first
step into the world of Free Software. Once they joined online, they would be
mentored on simple tasks and finally become full-fledged contributors! Of
course, universities would love the program, and with some luck start to
participate more actively, giving their students assignments which result in
code being written for the project, and much more.

\subsection*{\dots\ which didn't work\dots}
I have seen the idea from the fictitious story above in many forms in many
different communities and projects. It is a great idea and could be very
powerful! We all know you have to start early -- our proprietary competition is
pretty darn good at this. We also know we have arguments enough to convince
universities and students to participate -- FOSS is the future, it provides
great skill development opportunities, skills in Linux programming or
administration are in higher demand than another Java or .NET developer or
Windows sysadmin and most importantly: it is more fun. Somehow, however, if you go
to universities, you do not see many posters inviting you to join Free Software
projects. Most professors have never heard of it. What went wrong? Let me
continue the story.

\subsection*{\dots\ not because lack of effort\dots}
The team had a long discussion about it. First brainstorm style -- many ideas on
how to realize the idea came in. The team leader collected the work and put it
on the wiki. A plan was made with a time line and the team leader appointed
people responsible for certain parts. Some started writing  course materials,
others looked up university contact information and put it in a list. They asked
frequently for input and ideas on the mailing list and got plenty of responses
for further course material, which the leader added to the list of things to
write. It all had to be done in the free time of the volunteers, but you could
always count on the leader to remind volunteers of the schedule.

After a few months a structure was visible and many pages in the wiki were
created. Meanwhile, however, the number of people involved decreased from the
initial discussion with over 30 to about 5 still soldiering on. The leader
decided to revise the road map with proposed deadlines and after a few calls on
the mailing list 10 new volunteers committed to a variety of tasks. The pace
picked up a bit again. Quite a bit of what had been done before had to be
updated and there were other adjustments needed. Unfortunately, things kept
slipping and the number of people doing things kept decreasing. Monthly sprints
were introduced which did indeed result in some more work being finished. But
there was simply too much to do. After about a year, the last people gave up. A
stale wiki page and some outdated materials are all that is left\dots

\subsection*{\dots\ but because it was too ambitious.}
So why did it not work? The team did everything according to the best project
management practices you will find on the web\dots\ brainstorming, then creating
a plan, time lines, clear goals and responsibilities\dots\ They did the right
volunteer things: ask people, engage them, give everyone an opportunity to voice
his/her opinion. It should have worked!

It did not, because of a simple reason: it was too ambitious. It is a trend.
Amazing ideas receive lots of comments, get written down in great plans which
result in incomplete wiki pages leading to too little implementation finally
fading into nothingness.

Leaders have to recognize that how a team works in FOSS is not the same as in a
structured, managed environment like a company. People tend to be around when
there is something exciting, like a big release, and then disappear until the
next exciting thing. Creating a community team should never assume that the
people will stay fully committed the entire length of time. You have to factor
in that they will be in for a while and then disappear for longer periods and
then come back. The leaving and joining creates a lot of overhead so that little
gets done. Yes, we can lead people, but we cannot manage people, and once you
learn to give up the management aspect, you can focus more on things you need to
do in the immediate short term.

So instead of planning big things, find something small, doable and useful in
itself. Not a wiki page with a plan, but the first step of what you aim for. And
then, lead by doing. Make a rough first draft of an article. Make a first
version of a folder. Copy-paste from whatever exists, or improve something which
was already available. Then present the result, drafty as it is, to the team and
ask if someone wants to make it better. Do something small and it will work.

\subsection*{Don't plan, just do\dots}
So how do you do something as big as the university student plan? You don't! At
least, not directly. Discussing this with the whole team, planning -- it will
surely make for a fun discussion which can last weeks. But it will not get you
far. Instead, keep the plan to yourself. Seriously.

I am not saying that you should not talk about it -- you can. Share the ambition
with whoever is interested. And it is OK if they give suggestions. But do not
rely on it, do not make plans which go much further than the first 1-2 steps.
Instead, execute. Build on what is there. Send a draft of a new or improved
flyer to the mailing list. Ask someone who gave a course on your project to
share the material and improve it a bit. Those whose work you build on might
help you out! The people you spoke with about the plan who share your vision
might help you too. This way, you will frequently finish something -- a flyer,
an improved website, a presentation to be used. And people can, slowly, start
using it. Ambassadors can go to their local universities, using a few of the
things you have already created. To do what they do, they surely have to create
some missing materials -- which can go on the wiki as well. And you make
progress.

\subsection*{\dots\ and get your pie in the sky!}
In community marketing, strategy is not on the wiki. It is not in a plan nor a
time line. Neither is it discussed every week with the whole team. It is part of
a vision which has grown over time. It is carried by a few central people and
inspires the short-term plans and objectives. And it is shared by the team. But
it has no time line and it can not fail. It is flexible and does not depend on
anything or anyone in particular. And it will always be a pie in the sky\dots

So if you want to lead in a Free Software community marketing effort, keep that
big picture a big picture. Do not plan too much, but get things done!
